---
title: "Parameter Estimation for the PABM"
output: pdf_document
# output: html_document
# output:
#   pdf_document:
#     citation_package: natbib
#     number_sections: yes
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e}
- \setcitestyle{numbers,square,comma}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r echo = FALSE}
library(mclust)
library(magrittr)
library(ggplot2)
```

First, we note that the $\lambda^{(kl)}$ are only identifiable up to multiplicative constant. Instead, we aim to identify each block $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 

Let $M^{(kl)} = \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$. 
We can consider $M^{(kl)}$ as the edge probability matrix of a GRDPG with 
latent configuration 
$X = \frac{1}{2} \begin{bmatrix} \lambda^{(kl)} & \lambda^{(kl)} \\ 
\lambda^{(lk)} & -\lambda^{(lk)} \end{bmatrix}$ and signature $(1, 1)$. 

Treating $M^{(kl)}$ as an edge probability matrix, draw 
$\hat{M}^{(kl)} \sim BernoulliGraph(M^{(kl)})$, and let $\hat{X}$ be its ASE. 
Then $\hat{M}^{(kl)} = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$ and  $\|\hat{X} Q - X\|_{2 \to \infty} = O_P(\frac{(\log n_{kl})^c}{\sqrt{n_{kl}}})$ where $n_{kl} = n_{lk} = n_k + n_l$.

Then we get $\|\hat{M}^{(kl)} - M^{(kl)}\|$ 
$= \|(\hat{X} Q - X) I_{1,1} X^\top + X I_{1,1} (\hat{X} Q - X)^\top + (\hat{X} Q - X) I_{1,1} (\hat{X} Q - X)^\top\|$  
$\leq 2 \|\hat{X} Q - X\| \|I_{1, 1}\| \|X\| + \|I_{1,1}\| \|\hat{X} Q - X\|^2$  
$= O(\frac{(\log n_{kl})^c}{\sqrt{n_{kl}}})$